Optimizing Structural Properties Of Neural Networks With Genetic Algorithms

### 2015-08-03 paper/paper.tex
planning

    \newcommand{\param}[1]{\texttt{#1}}
    \newcommand{\feedback}{\param{feedback}}
    \newcommand{\gain}{\param{gain}}
    \newcommand{\size}{N}
    \newcommand{\ERprob}{p}
    \newcommand{\RINGk}{k}
    \newcommand{\FFlayers}{l}

@@

Using the above questions as a guideline, we decided to analyze what
network characteristics make the FORCE algorithm, introduced in
\citep{FORCE}, especially suitable for learning periodic patterns.  More
precisely, we wanted to identify which properties of a network are
important for its learning success, and how to choose their
values for an optimal result.  To this end, one option would be to take an
initial network and an adequate periodic pattern to be learned, and
optimize the network's internals for this very task. Then patterns may
emerge that hint at how an optimal network design looks.

@@

To address these issues we looked at \emph{parameterized random networks}, or
in the language of genetics, at different \emph{network genotypes}. These
genotypes contain parametric information about the networks (\emph{phenotypes}) to be created. 
Different phenotypes with the same genotype may thus
have strong quantitative differences, but they share structural qualities
whose fitness for a preassigned task we wanted to analyze and improve.

Since calculating the fitness of a genotype required the evaluation of many
different phenotypes and because of the resulting stochastic nature of the
problems we faced, we decided to use genetic algorithms for the
optimization process.

\section{Concepts and Model}
\paragraph{Network Model}

MAYBE WE SHOULD MOTIVATE FORCE HERE, WHICH THEN NATURALLY LEADS TO THE CHOICE OF NETWORK BELOW.

We chose the same rate based network architecture
that is used in the original FORCE publication \citep{FORCE} (denoted as
architecture A) and also in \citep{RM}, where our notation mainly stems
from. So we look at a network $\mathcal{N}$ with $N$ internal neurons and
states $x_i$ for $1 \le i \le N$ that loosely represent the neurons'
membrane potential. The firing rate $r_i$ of the $i$-th internal neuron is
given by $r_i = \tanh x_i$.  Further, the network contains $R$ readout
neurons with states $z_i$ for $1 \le i \le R$, where
\begin{equation} 
    z_i = \sum_{j=1}^{N}\omega\Ix{read}_{ij}\,r_j~.
\end{equation}
Taking into account both internal dynamics as well as an external feedback
pathway, the dynamics $t \mapsto x_i(t)$ of the single neurons are governed
by \footnote{~Note that we did not include external input signals in this
formula, as is done in \citep{FORCE} and \citep{RM}. We in fact enabled
inputs in our code base, but did not use them for the main results.}
\begin{equation}
    \dot{x}_i(t) = -x_i(t) + \sum_{i=1}^N \omega\Ix{rec}_{ij}\, r_j(t) + 
                   \sum_{i=1}^R \omega\Ix{fb}_{ij}\, z_j(t)~,
\end{equation}
where we introduce the internal recurrent synapse weights
$\omega\Ix{rec}_{ij}$ and the feedback synapse weights
$\omega\Ix{fb}_{ij}$. One can see that the neurons exponentially
decay if they are not presented stimulus by other neurons. 
We solved this system of differential equations by
a simple Newton method with integration time steps $dt = 0.01$, which was
also used in \citep{FORCE}. (How we constructed $\omega\Ix{rec}$ and
$\omega\Ix{fb}$ will be discussed in \emph{Parametric Random Networks})

@@

\paragraph{Parametric Random Networks -- Genotypes} Since we are not
interested in single network properties but rather want to find out which
structural elements of networks are important for learning success, we
introduced \emph{parametric random networks}, or \emph{network genotypes}.
Such a genotype $G$ consists of a set $\Theta$ of parameters that determine
which concrete networks are created when $G$ is expressed in a single
phenotype. Technically $G$ may be understood as a stochastic network
generator that returns a network $\mathcal{N}=G(\omega)$ for every seed
value $\omega$.
The genotypes' parameters $\theta\in\Theta$ were either integer or floating
point values and can be classified as affecting either (1) the topology of
$\omega\Ix{rec}$ or (2) the weight values of $\omega\Ix{rec}$ and
$\omega\Ix{fb}$. Important examples for (1) are the networks size $N$, the
occupation probability $p$ (for random Erdös-Renyi networks), the
neighborhood range $k$ (for ring topologies), or the number $l$ of layers
(for feed-forward networks). We furthermore introduced ratio parameters
that allowed us to interpolate between different topologies.  Parameters of
type (2) were the feedback strength $\texttt{feedback}$ (scaling of
$\omega\Ix{fb}$) and the gain value \gain (scaling of
$\omega\Ix{rec}$ by ).

@@

\section{Methods and Implementation}

\paragraph{Tasks}
For the learning to work properly, the problems presented to the
networks have (for the most part) to be solvable by FORCE. While
it was shown \citep{FORCE} that large recurrent neural networks are capable of
learning very complex functions, the optimization requires a large number
of learnings. Therefore, the networks have to be restricted to being quite
small ($\approx 100$ neurons) and are consequently less powerful.

To get a simple criterion to estimate whether it might be possible to learn a given function 
using FORCE with 100 neurons, we look at a single sinusoidal of frequency $\omega$ and 
determine the range of $\omega$ in which an (unoptimized) network (i.e. \gain, 
\feedback, \ERprob) is able to learn the function (see fig. \ref{}).

Since we expect/hope that the optimized network extends this range of possible frequencies,
we choose the frequencies for the optimization tasks from the slightly larger interval 
$[0.6e-2, 4]$ such that their logarithms are uniformly spaced. This allows for good resolution 
for both high and low frequencies.

\paragraph{Success}
In order to optimize the function reconstruction of a force network,
a measure $f(r', r)$ for the concordance of the target function
$r'(t)$ and the network readout $r(t)$ is necessary. Simply using the
difference $f\ix{d}(r, r') = \left\|r-r'\right\|_2$ has the
drawback that small phase differences could produce huge differences in
quality, even though $r'$ has the exact same shape as $r$.

A measure that is time-shift independent and maximized for functions of identical shape is the 
maximum of the cross correlation 
\begin{align}
    f\ix{c}(r, r') = \max_t \frac{\left<r, r'(\cdot - t)\right>}{\sqrt{\left< r, r\right> \left<r', r' \right>}}~.
\end{align}
Algorithmically, the signals are split into overlapping chunks which are correlated separately. 
This saves computational power and prevents minute differences in frequency to accumulate into 
a huge phase shift, i.e. the phase only has to remain relatively
constant over the timescale of a single chunk (about 10 seconds of simulated time).

@@

\paragraph{Genetic Optimization}
Due to the stochastic nature of the fitness function $F_C(G)$ of a generator $G$, and the fact that single evaluations
of $F_C$ are computationally expensive, thus preventing increasing the number of samples such that the estimation $E(F_C)$
becomes sufficiently smooth, typical optimization methods like gradient descent or simulated annealing are not usable.

 On approach would be to use techniques from Monte Carlo Integration, e.g. stratified or importance sampling, to aid in the
evaluation of $E(F_C)$, but since $G(\lambda)$ depends non-continuously on $\lambda$, it can only be applied to the dependence 
on the challenge $C$. This would mitigate the effects, but not solve the problem.

Therefore, we optimized using a genetic algorithm, which is inherently capable of coping with the stochastic fitness function.

In each iteration of the algorithm, a generation (set) of $N$ individuals (Generators) $G_t = \left\{ g_i \right\}$ is converted into
a new one $G_{t+1}$ according to the following scheme:
\begin{description}
 \item[Evaluation] The fitness $f_i = F_C(g_i)$ of each individual is measured using at least 20 samples (and adaptively more if the $f_i$ are spaced closely, to at msot 100).
 \item[Selection] All individuals are grouped into disjunct pairs for which we check which one has the higher $f$. This is repeated $K=10$ times, counting the wins of each individual. Based thereon we use the $p=50\%$ best individuals $S$ to build the next generation.

By using pair comparisons instead of simple ordering according to $f$, we allow weaker individuals to continue into the next generation albeit with low probability), keeping the gene pool more diverse.

\item[Reproduction]
We generate the next generation $G_{t+1}$ by first generating $2N$ candidates by \emph{mutating} or \emph{recombining} elements of $S$, i.e. either changing a single parameter randomly or taking two individuals and randomly choose for each parameter from which parent to take the value. 
We then perform a single sample fitness test and remove all networks that fail, until only $N$ networks are left which form $G_{t+1}$. 
This way, the most unpromising elements of the new generation are removed before the expensive fitness evaluation. 
\end{description}

\paragraph{Choice of Parameters}

\section{Results}

>>>>

@@

Using the above questions as a guideline, we decided to analyze what
network characteristics makes the FORCE algorithm, introduced in
\citep{FORCE}, especially suitable for learning periodic patterns.  More
precisely, we wanted to identify which properties of a network are
important for its learning success, and how to choose these properties'
values for an optimal result.  To this end, one option would be to take an
initial network and an adequate periodic pattern to be learned, and to
optimize the network's internals for this very task. Then patterns may
emerge that hint at how an optimal network design looks like.

@@

To address these issues we looked at \emph{parameterized random networks}, or
in the language of genetics, at different \emph{network genotypes}. These
genotypes contain parametric information about the networks to be created,
the \emph{phenotypes}. Different phenotypes with the same genotype may thus
have strong quantitative differences, but they share structural qualities
whose fitness for a preassigned task we wanted to analyze and improve.

Since calculating the fitness of a genotype required the evaluation of many
different phenotypes and because of the resulting stochastic nature of the
problems we faced, we decided to use genetic algorithms for the
optimization process. Other more direct optimization methods like simulated
annealing or gradient descent seemed unfeasible for this.

\section{Concepts and Methods}
\paragraph{Network Model} We chose the same rate based network architecture
that is used in the original FORCE publication \citep{FORCE} (denoted as
architecture A) and also in \citep{RM}, where our notation mainly stems
from. So we look at a network $\mathcal{N}$ with $N$ internal neurons and
states $x_i$ for $1 \le i \le N$ that loosely represent the neurons
membrane potential. The firing rate $r_i$ of the $i$-th internal neuron is
given by $r_i = \tanh x_i$.  Further, the network contains $R$ readout
neurons with states $z_i$ for $1 \le i \le R$, where
\begin{equation} 
    z_i = \sum_{j=1}^{N}\omega\Ix{read}_{ij}\,r_j~.
\end{equation}
Taking into account both internal dynamics as well as an external feedback
pathway, the dynamics $t \mapsto x_i(t)$ of the single neurons are governed
by \footnote{~Note that we did not include external input signals in this
formula, as is done in \citep{FORCE} and \citep{RM}. We in fact enabled
inputs in our code base, but did not use them for the main results.}
\begin{equation}
    \dot{x}_i(t) = -x_i(t) + \sum_{i=1}^N \omega\Ix{rec}_{ij}\, r_j(t) + 
                   \sum_{i=1}^R \omega\Ix{fb}_{ij}\, z_j(t)~,
\end{equation}
where we introduce the internal recurrent synapse weights
$\omega\Ix{rec}_{ij}$ and the feedback synapse weights
$\omega\Ix{fb}_{ij}$. One can see that the neurons exponentially
decay if they are not presented stimulus by other neurons. 
We solved this system of differential equations by
a simple Newton method with integration time steps $dt = 0.01$, which was
also used in \citep{FORCE}. (How we constructed $\omega\Ix{rec}$ and
$\omega\Ix{fb}$ will be discussed in \emph{Parametric Random Networks})

@@

\paragraph{Parametric Random Networks -- Genotypes} Since we are not
interested in single network properties but rather want to find out which
structural elements of networks are important for learning success, we
introduced \emph{parametric random networks}, or \emph{network genotypes}.
Such a genotype $G$ consists of a set $\Theta$ of parameters that determine
which concrete networks are created when $G$ is expressed in a single
phenotype. Technically $G$ may be understood as a stochastic network
generator that returns a network $\mathcal{N}=G(\omega)$ for every seed
value $\omega$.
The genotypes' parameters $\theta\in\Theta$ were either integer or floating
point values and can be classified as affecting either (1) the topology of
$\omega\Ix{rec}$ or (2) the weight values of $\omega\Ix{rec}$ and
$\omega\Ix{fb}$. Important examples for (1) are the networks size $N$, the
occupation probability $p$ (for random Erdös-Renyi networks), the
neighborhood range $k$ (for ring topologies), or the number $l$ of layers
(for feed-forward networks). We furthermore introduced ratio parameters
that allowed us to interpolate between different topologies.  Parameters of
type (2) were the feedback strength \texttt{feedback} (scaling of
$\omega\Ix{fb}$) and the gain value \texttt{gain} (scaling of
$\omega\Ix{rec}$ by ).

@@

\paragraph{Genetic Optimization}
Genetic Optimization:

- Hint / vague description of the codebase

\paragraph{Tasks and Challenges}
For the optimization to work properly, the problems presented to the networks 
have (for the most part) to solvable by FORCE learning. While it was shown \citep{}
that large recurrent neural networks are capable of learning very complex functions
using the FORCE algorithm, the optimization requires a large number of learnings.
Therefore, the networks have to be restricted to being quite small ($\approx 100$ neurons)
and as such are less powerful.

To get a simple criterion to estimate whether it might be possible to learn a given function 
using FORCE with 100 neurons, we look at a single sinusoidal of frequency $\omega$ and 
determine the range of $\omega$ in which an (unoptimized) network (i.e. GAIN, 
FEEDBACK, p, Erdös Renyi) is able to learn the function (see fig. \ref{}).

Since we expect/hope that the optimized network extends this range of possible frequencies,
we choose the frequencies for the optimiztion tasks from the slighly larger interval 
$[0.6e-2, 4]$ such that their logarithms are uniformly spaced. This allows for good resolution 
for both high and low frequencies.

\paragraph{Assessing the quality of reconstructed functions}
In order to optimize the function reconstruction of a force network, a measure $f(g, g^\prime)$ for
the concordance of target $g$ and network output $g^\prime$ is necessary. Simply using the difference
$f_d(g, g^\prime) = \left\| g-g^\prime \right\|_2$ has the drawback that small phase differences could 
produce huge differences in quality, even though $g^\prime$ has the exact same shape as $g$.

A measure that is timeshift independent and maximized for functions of identical shape is the 
maximum of the cross correlation 
\begin{align}
f_c(g, g^\prime) = \max_t \left<g, g^\prime(\cdot - t)\right> / \sqrt{\left< g, g\right> \left<g^\prime, g^\prime \right>}.
\end{align}
Algorithmically, the signals are split into overlapping chunks which are correlated separately. 
This saves computational power and prevents minute differences in frequency to accumulate into 
a huge phase shift, i.e. the phase only has to remain relatively
constant over the timescale of a single chunk (i.e. about 10 seconds of simulated time).

@@

\section{Results}
Baseline
gain and feedback tests

(feedback shifts possible frequencies to higher values)

success of omega: erdös-renyi, ring
optimizations: ER topology, Ring Topology, Mixing
